---
title: "R Tutorial - Documentation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Congrats! You've made it to the end of the R workshop! We have prepared this documentation for you, where you will find all the code we wrote during the workshop. Don't hesitate to go through it again at your own pace to make sure you understand everything ;) 

Have fun with R!

Celes, Lander, Lisa, Mustafa and Nikolas

## Celes part

## R Data Types

R has 5 basic data types that are :  
* character  
* numeric  
* integer  
* complex  
* logical  

**Character**  
A character is a text (or string) value typed between single ('') or double ("") quotes. 

```{r}
#To assign a value to a variable, you need to use the command: variable <- value
my_character <- "hello there"

#To print out the data type of a variable, you can use the function typeof() or class()
typeof(my_character)
```
**Numeric**  
A numeric is a decimal value (also called double-precision real value). 

```{r}
my_numeric <- 3.14
typeof(my_numeric)
```
**Integer**  
Integers are a subset of the numeric data type. Integers are whole values and are specified by the letter "L".

```{r}
my_integer <- 24
typeof(my_integer)
```
```{r}
my_integer <- 24L
typeof(my_integer)
```
**Complex**  
A complex number is a value with real and imaginary parts.

```{r}
my_complex <- 2+5i
typeof(my_complex)
```
**Logical**  
A logical is a boolean value (*TRUE* or *FALSE*). Be careful: Logicals are case sensitive. Writing *True* or *true* does not work.

```{r}
my_logical <- TRUE
typeof(my_logical)
```
## Arithmetic with R
R can also be used as a calculator. R follows the standard order of priority and accepts the following operators:  
Addition: `+`  
Subtraction: `-`  
Multiplication: `*`  
Division: `/`  
Exponentiation: `^`  
  
Missing values are  represented  by  `NA`  (Not  Available)  without  quotes.  `NA`  represents  both  missing character and numeric data. Impossible values (e.g., dividing by zero) are represented by the symbol `NaN` (Not A Number)

```{r}
x <- 4+2*6
y <- 2^10
    
#To print out the value of a variable, you can use the function print() or just type the name of the variable
print(x)
y
```
## R Data Structures
R’s basic data structures include the vector, list, matrix, data frame, and factors. Some of these structures require that all members be of the same data type (e.g. vectors, matrices) while others permit multiple data types (e.g., lists, data frames).

**Vectors**  
A vector is the simplest and most used data structure in R. It is a collection of data values of the same type (e.g., all numerics, characters, logicals...    
The easiest way to create a vector is to use the c() function that *concatenates* several given elements.  

```{r}
x <- c("blue", "green", "orange", "red", "yellow") #vector of characters
y <- c(44, 33, 22, 11) #vector of numerics
z <- c(TRUE, FALSE, FALSE) #vector of logicals
```

Let's have a look at the attributes of the vectors we just created:  

```{r}
class(x) #to print the type of elements in the vector
length(y) #to print the length of the vector
str(z) #to print the structure/content of the vector
```
When combining vectors of different data types, one data type will outrule the other one(s).  
The rule is: Character > Numeric > Integer > Logical.

```{r}
i <- c(x, y)
print(i)
class(i)
```


Manipulation of vectors: adding elements, retrieving elements, change data type...  

**Adding an element**

```{r}
x <- c(x, "white")
x

y <- c(55, y)
y
```


```{r}
getwd()
summary(iris)
```
